---
title: "Exploring Spatio-Temporal Data Lab2_3"
author: "Erick Calderón-Morales"
date: "February 7, 2019"
output: 
  html_document:
    code_folding: show
    fig_height: 30
    fig_width: 30
    keep_md: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '2'
  word_document:
    toc: yes
    toc_depth: '2'
---

#Lab 2.3: Exploratory Data Analysis

```{r message=FALSE, warning=FALSE}
library(CCA)
library(tidyverse)
library(gstat)
library(sp)
library(spacetime)
library(STRbook)

set.seed(1)
```

In this Lab we consider the construction and visualization of the empirical means and covariances, the use of empirical orthogonal functions and their associated principal component time series, semivariogram analysis, and spatio-temporal canonical correlation analysis.


```{r}
#Data
data("NOAA_df_1990", package = "STRbook")

tmax <- filter(NOAA_df_1990,
               proc == "Tmax" &
               month %in% 5:9 &
               year == 1993)
tmax$t <- tmax$julian - 728049

```

#Empirical Spatial Means

```{r}
head(tmax)
```


```{r}
            # group by lon-lat
spat_av <- group_by(tmax,lat, lon) %>% 
  
           # mean for each lon-lat
           summarise(mu_emp = mean(z))

head(spat_av)
```


```{r}
#plot average maximun temperature per station 

#latitude plot
(lat_means <- ggplot(spat_av)+
             geom_point(aes(lat, mu_emp))+
             theme_bw())
```


```{r}
(lon_means <- ggplot(spat_av)+
             geom_point(aes(lon, mu_emp))+
             theme_bw())
```



#Empirical Temporal Means

```{r}
tmax_av <- group_by(tmax,date) %>% 
           summarise(meanTmax= mean(z))
```


```{r}
(gTmaxav <- ggplot()+
   #plot de lineas
  geom_line(data = tmax, aes(x=date, y=z, group= id),
            colour= "blue", alpha= 0.4) +
   #agregar una linea media
   geom_line(data= tmax_av, aes(x= date, y= meanTmax))+
   theme_bw())
```



#Empirical Covariances

It is important that all trends are removed, one way to do this is to first fit a linear model 
that has spatial and/or temporal covariates. Then plot the empirical covariances of the detrended
data (i.e., the residuals).

We observed a quadratic tendency of temperature over the chosen time span. 


##First fit a linear model 

```{r}
lm1 <- lm(z ~ lat+ t + I(t^2), data = tmax )

#Store the residuals
tmax$residuals <- residuals(lm1)
```



```{r}
#Consider the spatial locations
spat_df <- filter(tmax, t==1) %>% 
  #extraer solo lon lat 
  select(lon, lat) %>% 
  
  #ordenar lon lat
  arrange(lon, lat)

#Number of stations
m <- nrow(spat_av)
```


##Compute the empirical covariance matrix

__Ver video para entender matrices de covarianza__
https://www.youtube.com/watch?v=9B5vEVjH2Pk
```{r}
#First put the data into space-wide format

x <- select(tmax, lon, lat, residuals, t) %>% 
  
  #ponga t en las columnas y muestre los valores de los residuos 
  spread(t, residuals) %>% 
  
  #eliminar lon lat 
  select(-lon, -lat) %>% 

  #make space-wide matrix
  t()
  

```

```{r}
#Compute covariance matrix
lag0_cov <- cov(x, use = "complete.obs")

#covariance between the residuals from X excluding the first time point and X
#excluding the last time point
lag1_cov <- cov(x[-1,], x[-nrow(x),], use = "complete.obs")

```


With a roughly square domain, split the domain into either latitudinal or longitudinal strips



```{r}
spat_df$n <- 1:nrow(spat_df)
lim_lon <- range(spat_df$lon)

#crear 4 long strip boundaries
lon_strips <- seq(lim_lon[1],
                  lim_lon[2],
                  length=5)

spat_df$lon_strip <- cut(spat_df$lon,
                         lon_strips,
                         labels = FALSE,
                         include.lowest = T)  

#Now that we know in which strip each station falls  
head(spat_df)
```


The empirical spatial covariance matrices reveal the presence of spatial correlation in the residuals.

```{r}
par(mfrow=c(2,2))
plot_cov_strips(lag0_cov, spat_df )
```
```{r}
par(mfrow=c(2,2))
plot_cov_strips(lag1_cov, spat_df )
```


#Semivariogram Analysis

Computing the empirical semivariogram is much faster when
using objects of class __STFDF__ rather than STIDF since the regular space-time structure can
be exploited.


```{r}
data("STObj3", package= "STRbook")
STObj4 <- STObj3[, "1993-07-01::1993-07-31"]

```


```{r}
vv <- variogram(object = z~ 1 +lat,
                
                #July data
               data = STObj4,
               
               #spatial bin (80km)
               width=80,
               
               #consider points <1000km apart
               cutoff =1000,
               
               #0 days to 6 days
               tlags= 0.01:6.01)
```


The plot suggests that there are considerable
spatio-temporal correlations in the data; spatio-temporal modeling of the residuals is
thus warranted.
```{r}
plot(vv)
```



#Empirical Orthogonal Functions

Empirical orthogonal functions (EOFs) can reveal spatial structure in the data and can also
be used for subsequent dimensionality reduction

```{r}
data("SSTlandmask", package = "STRbook")
data("SSTlonlat", package = "STRbook")
data("SSTdata", package = "STRbook")


delete_rows <- which(SSTlandmask == 1)
SSTdata <- SSTdata[-delete_rows, 1:396]
```





```{r}
## Put data into space-wide form
Z <- t(SSTdata)
dim(Z)
```


```{r}
## First find the matrix we need to subtract:
spat_mean <- apply(SSTdata, 1, mean)
nT <- ncol(SSTdata)

## Then subtract and standardize:
Zspat_detrend <- Z - outer(rep(1, nT), spat_mean)
Zt <- 1/sqrt(nT - 1)*Zspat_detrend

#The matrix V contains the EOFs in space-wide format. We change the column
#names of this matrix, and append the lon-lat coordinates to it as follows.
E <- svd(Zt)

V <- E$v

colnames(E$v) <- paste0("EOF", 1:ncol(SSTdata)) # label columns
EOFs <- cbind(SSTlonlat[-delete_rows, ], E$v)
head(EOFs[, 1:6])

```

```{r}
TS <- data.frame(E$u) %>% # convert U to data frame
mutate(t = 1:nrow(E$u)) %>% # add a time field
gather(EOF, PC, -t) # put columns (except time)
                    # into long-table format with
                   # EOF-PC as key-value pair
```

```{r}
TS$nPC <- TS$PC * sqrt(nT-1)
```



```{r}
ggplot(EOFs) + geom_tile(aes(x = lon, y = lat, fill = EOF2)) +
fill_scale(name = "degC") + theme_bw() +
xlab("Longitude (deg)") + ylab("Latitude (deg)")
```

#Spatio-Temporal Canonical Correlation Analysis

```{r}
nEOF <- 10
EOFset1 <- E$u[1:(nT-7), 1:nEOF] * sqrt(nT - 1)
EOFset2 <- E$u[8:nT, 1:nEOF] * sqrt(nT - 1)
```

```{r}
cc <- cancor(EOFset1, EOFset2) # compute CCA
options(digits = 3) # print to three d.p.
print(cc$cor[1:5]) # print
```

```{r}
print(cc$cor[6:10])
```

```{r}
CCA_df <- data.frame(t = 1:(nT - 7),
CCAvar1 = (EOFset1 %*% cc$xcoef[,1])[,1],
CCAvar2 = (EOFset2 %*% cc$ycoef[,1])[,1])
```

```{r}
t_breaks <- seq(1, nT, by = 60) # breaks for x-labels
year_breaks <- seq(1970,2002,by=5) # labels for x-axis

(g <- ggplot(CCA_df) +
geom_line(aes(t, CCAvar1), col = "dark blue") +
geom_line(aes(t, CCAvar2), col = "dark red") +
scale_x_continuous(breaks = t_breaks, labels = year_breaks) +
ylab("CCA variables") + xlab("Year") + theme_bw())
```

